MCMC convergence diagnostics

### 2025-08-05 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Run through of MCMC convergence report with new results

```{r}
# This is code I tried when one chain was not converging -- trying for a rerun instead!
remove_chains = function(fit, drop) {
  sim <- fit@sim
  sim$samples <- sim$samples[-drop]
  sim$chains <- sim$chains - length(drop)
  sim$warmup2 <- sim$warmup2[-drop]
  sim$n_save <- sim$n_save[-drop]
  fit@sim <- sim
  fit
}
```

@@

ggplot(rhats_df, aes(x = value, y = parameter)) +
  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7, col = "#56B4E9") +
  lims(x = c(NA, 1.03)) +
  labs(x = "Potential scale reduction factor", y = "Parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

@@

```{r}
ess_histogram <- data.frame(mcmc_summary) %>%
  tibble::rownames_to_column("param") %>%
  ggplot(aes(x = n_eff)) +
    geom_histogram(col = "grey60", fill = "grey80") +
    labs(x = "Effective sample size", y = "Count", tag = "B")

@@

```{r}
worst_eff <- bayesplot::mcmc_trace(mcmc, pars = names(which.min(mcmc_summary[, "n_eff"]))) + labs(tag = "A") + guides(col = "none")
worst_rhat <- bayesplot::mcmc_trace(mcmc, pars = names(which.max(rhats))) + labs(tag = "B")

worst_eff + worst_rhat

ggsave("worst-trace.png", h = 3, w = 6.25)
```

@@

pairs_df <- extract(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a")) %>%
  as.data.frame()

@@

nb <- area_merged %>%
  filter(area_level == max(area_level)) %>%
  bsae::sf_to_nb()

>>>>

@@

ggplot(rhats_df, aes(x = value, y = parameter)) +
  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7, col = "#56B4E9") +
  lims(x = c(NA, 1.012)) +
  labs(x = "Potential scale reduction factor", y = "Parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

@@

```{r}
ess_histogram <- data.frame(mcmc_summary) %>%
  tibble::rownames_to_column("param") %>%
  ggplot(aes(x = n_eff)) +
    geom_histogram(col = "grey60", fill = "grey80") +
    scale_x_log10() +
    labs(x = "Effective sample size (log 10 scale)", y = "Count", tag = "B")

@@

```{r}
worst_ess_par <- names(which.min(mcmc_summary[, "n_eff"]))
worst_rhat_par <- names(which.max(rhats))

worst_df <- bayesplot::mcmc_trace_data(mcmc, pars = c(worst_ess_par, worst_rhat_par)) |>
  mutate(
    chain_group = case_when(
      chain %in% 1:6 ~ "Chains 1–6",
      chain %in% 7:12 ~ "Chains 7–12",
      chain %in% 13:18 ~ "Chains 13–18"
    ),
    chain_id = as.numeric(chain) %% 6 + 1
  )

worst <- ggplot(worst_df, aes(x = iteration, y = value, col = as.factor(chain_id))) +
  geom_line() +
  scale_color_manual(values = cbpalette) +
  facet_grid(chain_group ~ parameter) +
  guides(col = "none") +
  labs(x = "Iteration", y = "") +
  theme_minimal()

ggsave("worst-trace.png", worst, h = 4.5, w = 6.25)
```

@@

pairs_df <- rstan::extract(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a")) %>%
  as.data.frame()

@@

sf_to_nb <- function(sf){
  nb <- spdep::poly2nb(as(sf, "Spatial"))
  return(nb)
}

nb <- area_merged %>%
  filter(area_level == max(area_level)) %>%
  sf_to_nb()
